---
title: "Data visualisation in geoGraph"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data visualisation in geoGraph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 6)
options(digits = 4)
```

```{r loading libraries, include=FALSE}
library(geoGraph)
library(sf)
```

Manually editing graphs in *geoGraph*
=================================================

This vignette will cover the main functions for visualizing all objects in *geoGrapgh.* It first gives an overview of the basic plotting functions and then also has a section for more advanced visualisation using `ggplot2.`

## Basic visualizing

An essential aspect of spatial analysis lies in visualizing the data.
In *geoGraph*, the spatial grids (`gGraph`) and spatial data (`gData`) can be plotted
and browsed using a variety of functions.

### Plotting `gGraph` objects

Displaying a `gGraph` object is done through `plot` and `points` functions.
The first opens a new plotting region, while the second draws in the current plotting region;
functions have otherwise similar arguments (see `?plot.gGraph`).

By default, plotting a `gGraph` displays the grid of nodes overlaying a shapefile (by default,
the landmasses).
Edges can be plotted at the same time (argument `edges`), or added afterwards using `plotEdges`.
If the `gGraph` object possesses an adequately formed `meta$colors` component, the
colors of the nodes are chosen according to the node attributes and the color scheme specified in `meta$colors`.
Alternatively, the color of the nodes can be specified via the `col` argument in `plot`/`points`.

Here is an example using `worldgraph.10k`:

```{r wg10kdefplot, fig=TRUE, fig.width=8}
getColors(worldgraph.10k, res.type = "rules")
head(getNodesAttr(worldgraph.10k))
table(getNodesAttr(worldgraph.10k))
plot(worldgraph.10k, reset = TRUE)
title("Default plotting of worldgraph.10k")
```

 It may be worth noting that plotting `gGraph` objects involves plotting a fairly
large number of points and edges.
On some graphical devices, the resulting plotting can be slow.
For instance, one may want to disable `cairo` under linux: this graphical device yields better
graphics than `Xlib`, but at the expense of increased computational time.
To switch to `Xlib`, type:
```{r eval=FALSE}
X11.options(type = "Xlib")
```
and to revert to `cairo`, type:
```{r eval=FALSE}
X11.options(type = "cairo")
```

### Plotting `gData` objects

`gData` objects are by default plotted overlaying the corresponding `gGraph`.
To show this, we use the `cities` example from the vignette 'get started':
```{r citiesPlot2, fig=TRUE}
bordeaux <- c(-1, 45)
berlin <- c(13, 52)
baku <- c(44, 40)
timbuktu <- c(-3, 16)

cities.dat <- rbind.data.frame(bordeaux, berlin, baku, timbuktu)
colnames(cities.dat) <- c("lon", "lat")
row.names(cities.dat) <- c("Bordeaux", "Berlin", "Baku", "Timbuktu")
cities.dat$pop <- c(250000, 3500000, 2000000, 50000)
cities.dat

cities <- new("gData", coords = cities.dat[, 1:2], data = cities.dat[, 3, drop = FALSE], gGraph.name = "worldgraph.10k")
plot(cities, type = "both", reset = TRUE)
text(getCoords(cities), rownames(getData(cities)))
```

Note the argument `reset=TRUE`, which tells the plotting function to adapt the
plotting area to the geographic extent of the dataset.


To plot additional information, it can be useful to extract the spatial coordinates from the data.
This is achieved by `getCoords`.
This method takes an extra argument `original`, which is TRUE if original spatial coordinates
are sought, or FALSE for coordinates of the nodes on the grid.
We can use this to represent, for instance, the population sizes for the different cities:
```{r cities_plot, fig=TRUE}
transp <- function(col, alpha = .5) {
  res <- apply(
    col2rgb(col), 2,
    function(c) rgb(c[1] / 255, c[2] / 255, c[3] / 255, alpha)
  )
  return(res)
}

plot(cities, reset = TRUE)
par(xpd = TRUE)
text(getCoords(cities) + -.5, rownames(getData(cities)))
symbols(getCoords(cities)[, 1], getCoords(cities)[, 2],
  circ = sqrt(unlist(getData(cities))), inch = .2,
  bg = transp("red"), add = TRUE
)
```

