---
title: "Making custom grids in geoGraph"
author: "Dominik Jud"
date: "2026-02-20"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Making custom grids in geoGraph}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

Using discrete global grids in *geoGraph*
=================================================

This vignette illustrates how to create custom graphs in *geoGraph* using discrete global grids. This is especially useful when working over larger areas of the world, where a rectangular grid would introduce messy distortions. Using the `dggridR` package, we can create a custom hexagonal grids that provides a more accurate representation of the Earth's surface. We will then use this grid to create a `gGraph` object in *geoGraph* and assign environmental information to the nodes of the graph. This allows us to create a cost surface for connectivity analysis that takes into account both land/water classification and elevation variability.


```{r loading libraries, include=FALSE}
library(geoGraph) # graph analysis
library(elevatr) # for elevation data
library(sf)
library(terra)
library(tidyverse)
```

### Getting the elevation data

In the first step, we define the area of interest using a bounding box and then use the `get_elev_raster` function from the `elevatr` package to retrieve topographical and bathymetry data for that area. The resulting data is then converted to a `SpatRaster` object using the `rast` function from the `terra` package.

```{r getting topographic data}
# define the area of interest (Europe)
area_of_interest <- st_as_sfc(
  st_bbox(c(
    xmin = -25,
    xmax = 25,
    ymin = 34,
    ymax = 50
  ), crs = 4326)
)

aoi_bound <- st_sf(geometry = area_of_interest)

elevation_data <- get_elev_raster(
  locations = aoi_bound,
  z = 4,
  clip = "locations"
)

# convert to SpatRaster
spatr <- rast(elevation_data)
plot(spatr)
```

#### Makeing a land mask

Using this `SpatRaster` we can create a land mask by using the `make_land_mask` function from the `pastclim` package. This function identifies land and ocean areas based on the simple logic of moving the ocean up and down given the current relief profile for the time of choice. In this case, we use the present time (0 years before present) to create a land mask for our area of interest.

```{r making land mask}
# make land mask (everything above 0m is land)
land_mask <- make_land_mask(spatr, time_bp = 0) # using the present time
land_mask[is.na(land_mask)] <- 0 
plot(land_mask)
```

### Creating the graph object

Next, we create a new graph object using the `createNewGraph` function from the `geoGraph` package. We specify the geographic bounding box for for our area of interest and set the spacing between nodes in the graph. The resulting graph object represents a grid over the area of interest with cells of the about 16km spacing, and we can visualize the graph structure using the `plot` and `plotEdges` functions.

```{r creating graph object}
ggraph <- createNewGraph(geo_box = aoi_bound, spacing = 20)
plot(ggraph, reset = TRUE)
plotEdges(ggraph)
```

### Addiing the land/water mask to the graph

We then use the `assignRasterPoints` function to add the land/water mask to the graph. This function takes the graph object and the land mask raster as inputs and assigns the land/water values to the corresponding cells in the graph. In this example we classify each cell with land points in it as land by defining the 'fun' parameter in collapseNodeAttribute as 'min', which means that if there is at least one water point in the cell, the cell will be classified as water. We can then visualize the resulting graph with the land/water mask by plotting the graph and coloring the nodes based on the land/water values.

```{r}
water_graph <- assignRasterPoints(
  graph = ggraph,
  raster = land_mask,
  layer_name = "water"
)

# now we collapse the water attribute to get a land/sea classification for each node
water_land_graph <- collapseNodeAttribute(
  graph = water_graph,
  attribute = "water",
  fun = min,
  na.rm = TRUE
)

# add this to the habitat attribute
habitat <- factor(water_land_graph@nodes.attr$water,
  levels = c(0, 1),
  labels = c("sea", "land")
)

# reclassify the nodes in the gGraph object accordingly in the metadata
water_land_graph@nodes.attr$habitat <- habitat #TODO can we use a setAttribute here?

colors <- data.frame(
  habitat = c("sea", "land"),
  cost = c("blue", "green")
)

# change colors associated with the land meta info
water_land_graph@meta$colors <- colors

plot(water_land_graph, reset = TRUE)
```

#### Setting costs for the different habitat types

Finally, we can set the costs for moving through different habitat types based on their classification. Note that the cost here is arbitrary and can be adjusted based on the specific context of the analysis and the species or features being studied. Costs are inversely related to the ease of movement through a cell, so higher costs indicate more difficult terrain. For example, we can assign a cost of 10 for moving through sea cells and a cost of 1 for moving through land cells. If we now look at the least cost path between two cities (Madrid and Naples) using the `dijkstraBetween` function, we can see that the path tends to avoid sea cells and prefers to move through land cells.

```{r least cost path}
# set the costs for sea cells to 10 and land cells to 1
sea_cost <- 10
land_cost <- 1
water_land_graph@meta$costs <- data.frame(
  habitat = c("sea", "land"),
  cost = c(sea_cost, land_cost)
)

water_land_graph <- setCosts(water_land_graph, method = "mean", attr.name = "habitat")


point1 <- c(-3.7038, 40.4168) # Madrid
point2 <- c(14.2681, 40.8522) # Naples
cities.dat <- rbind.data.frame(point1, point2)
colnames(cities.dat) <- c("lon", "lat")
row.names(cities.dat) <- c("Madrid", "Naples")
cities <- new("gData", coords = cities.dat[, 1:2], gGraph.name = "water_land_graph")
path <- dijkstraBetween(cities)
plot(water_land_graph, reset = TRUE)
plot(path, col = "red", lwd = 2)
```

### Assigning elevation data to graph nodes

In the next step we can then use the `assignRasterPoints` function again to add elevation data to the nodes of the graph. In this case we need to chose how to incorporate the elevation data into the graph, since each node will have multiple raster points associated with it. So before using the collapseNodeAttribute function to get a single elevation value for each node, we can first have a look at the standard deviation of the elevation values for each node, which gives us an idea of the variability of elevation within each cell. This can be useful for identifying rugged areas, which typically have high variability in elevation and a higher cost of traveling through associated. For this we set the 'fun' parameter in the collapseNodeAttribute function to 'sd', which calculates the standard deviation of the elevation values for each node. We can then visualize the distribution of the standard deviation of elevation values for land cells by plotting a density plot.

```{r assigning elevation data}
elevation_graph <- assignRasterPoints(
  graph = water_land_graph,
  raster = spatr,
  layer_name = "elevation"
)

# now we collapse the elevation attribute to get the sd of the elevation for each node
elevation_graph <- collapseNodeAttribute(
  graph = elevation_graph,
  attribute = "elevation",
  fun = sd,
  na.rm = TRUE
)

# have a look at the distribution of sd_elevation values for land cells
sd_ele <- getNodesAttr(elevation_graph)%>%
  mutate(
    elevation = ifelse(habitat == "land", elevation, NA_real_)
  ) %>%
  pull(elevation)

# plot the density
plot(density(log(sd_ele), na.rm = TRUE),
  main = "Density of SD Elevation for Land Cells",
  xlab = "SD Elevation (m)", ylab = "Density"
)
```

### Reclassify the nodes into 'rugged' and 'very rugged'

Looking at the distribution of the standard deviation of elevation values for land cells, we can see that there are some cells with very high variability in elevation, which likely correspond to mountainous areas. We can then reclassify the nodes into 'rugged' and 'very rugged' based on an elevation threshold. Here we pick arbitrary thresholds for the standard deviation of elevation values to classify cells as 'rugged' and 'very rugged'. Cells with a standard deviation of elevation greater than 100 are classified as 'rugged', while those with a standard deviation greater than 150 are classified as 'very rugged'. We can then update the habitat attribute in the graph accordingly and visualize the resulting graph with the new habitat classifications.

```{r reclassifying nodes}
# set a threshold for rugged cells
rugged_threshold <- 150
very_rugged_threshold <- 300

## now we identify rugged and very rugged cells as those with sd_elevation > threshold
habitat_new <- case_when(
  is.na(sd_ele) ~ "sea",
  sd_ele > very_rugged_threshold ~ "very_rugged",
  sd_ele > rugged_threshold ~ "rugged",
  TRUE ~ "land"
)

# update the habitat attribute in the graph
# reclassify the nodes in the gGraph object accordingly in the metadata
elevation_graph@nodes.attr$habitat <- habitat_new


colors <- data.frame(
  habitat = c("sea", "land", "rugged", "very_rugged"),
  cost = c("blue", "green", "orange", "brown")
)

# change colors associated with the land meta info
elevation_graph@meta$colors <- colors

plot(elevation_graph, reset = TRUE)
```

Now if we now set the cost for rugged and very rugged cells to be higher than for land cells, we can see that the least cost paths between different land areas tend to avoid not only sea cells but also rugged and very rugged cells. This can be shown in the same toy example as above.

```{r least cost path with ruggedness}
# set the costs for sea cells to 10, land cells to 1, rugged cells to 3 and very rugged cells to 10
sea_cost <- 10
land_cost <- 1
rugged_cost <- 3
very_rugged_cost <- 10
elevation_graph@meta$costs <- data.frame(
  habitat = c("sea", "land", "rugged", "very_rugged"),
  cost = c(sea_cost, land_cost, rugged_cost, very_rugged_cost)
)
elevation_graph <- setCosts(elevation_graph, method = "mean", attr.name = "habitat")
point1 <- c(-3.7038, 40.4168) # Madrid
point2 <- c(14.2681, 40.8522) # Naples
cities.dat <- rbind.data.frame(point1, point2)
colnames(cities.dat) <- c("lon", "lat")
row.names(cities.dat) <- c("Madrid", "Naples")
cities <- new("gData", coords = cities.dat[, 1:2], gGraph.name = "elevation_graph")
path <- dijkstraBetween(cities)
plot(elevation_graph, reset = TRUE)
plot(path, col = "red", lwd = 2)
```

### Adding coastal cells

Now in a last step we can add a 'coast' category to the habitat classification for cells that are adjacent to any land based cells. This is done by checking the neighbors of each sea cell and reclassifying those that have at least one land or mountain neighbor as 'coast'. We can then visualize the resulting graph with the new habitat classifications, including the coastal cells.

```{r adding coastal cells}
# get the neighbor list from the graph
neigh_list <- elevation_graph@graph@edgeL
node_attr <- getNodesAttr(elevation_graph)

# get all nodes adjacent to land or mountain cells and reclassify them as "coast"
for (i_node in 1:nrow(node_attr)) {
  if (node_attr$habitat[i_node] == "sea") {
    neighbour_indices <- neigh_list[[i_node]]$edges # check structure!
    neighbour_land_values <- node_attr$habitat[neighbour_indices]
    if (any(neighbour_land_values %in% c("land", "rugged", "very_rugged"))) {
      # at least one land or mountain neighbour
      node_attr$habitat[i_node] <- "coast"
    }
  }
}

new_attribute <- node_attr

# create the full_graph
full_graph <- elevation_graph

# set the new node attributes
full_graph@nodes.attr <- new_attribute


colors <- data.frame(
  habitat = c("sea", "land", "rugged", "very_rugged", "coast"),
  cost = c("blue", "green", "orange", "brown", "lightblue")
)

# change the costs and colors associated with the land meta info
full_graph@meta$colors <- colors
plot(full_graph, reset = TRUE)
```

Now if we set the cost for coastal cells to be the same as for land cells, we can see that the least cost paths between different land areas tend to prefer crossing coastal cells over crossing rugged areas. 

```{r least cost path with coastal cells}
# set the costs for sea cells to 10, land cells to 1, rugged cells to 3, very rugged cells to 10 and coastal cells to 1
sea_cost <- 10
land_cost <- 1
rugged_cost <- 3
very_rugged_cost <- 10
coastal_cost <- 1
full_graph@meta$costs <- data.frame(
  habitat = c("sea", "land", "rugged", "very_rugged", "coast"),
  cost = c(sea_cost, land_cost, rugged_cost, very_rugged_cost, coastal_cost)
)
full_graph <- setCosts(full_graph, method = "mean", attr.name = "habitat")

point1 <- c(-3.7038, 40.4168) # Madrid
point2 <- c(14.2681, 40.8522) # Naples
cities.dat <- rbind.data.frame(point1, point2)
colnames(cities.dat) <- c("lon", "lat")
row.names(cities.dat) <- c("Madrid", "Naples")
cities <- new("gData", coords = cities.dat[, 1:2], gGraph.name = "full_graph")
path <- dijkstraBetween(cities)
plot(full_graph, reset = TRUE)
plot(path, col = "red", lwd = 2)
```

### Using different methods to calculate costs

So far the costs for moving through different habitat types were defined by the mean of the costs associated with the different habitat types. However, the `setCosts` function allows to use different methods to calculate the costs for each edge based on the node attributes. For example, instead of using the mean, we could specify a custom function that takes the maximum cost of the two nodes connected by an edge, which would mean that the cost of moving through an edge is determined by the more difficult habitat type of the two nodes. This can be done by defining a custom function `max.cost` and then using it in the `setCosts` function with the method set to "function". The resulting graph will have costs for each edge that reflect the maximum cost of the two nodes it connects.

```{r setting costs}
max.cost <- function(x1, x2) {
  pmax(x1, x2)
}

full_graph <- setCosts(
  full_graph,
  attr.name = "habitat",
  method = "function",
  FUN = max.cost
)

plot(full_graph, reset = TRUE)
plotEdges(full_graph)
```

### Combining costs

In many ecological applications, connectivity across landscapes depends on multiple environmental constraints rather than a single variable. In the previous sections we constructed terrain-based costs that capture differences in movement through sea, land, rugged, and very rugged areas. However, climatic factors can also influence movement and dispersal.
To illustrate how multiple environmental layers can be incorporated into a connectivity model, we now add an additional environmental gradient: `temperature`, which in this example is a random variable that we assign to the nodes of the graph. The movement cost between two nodes can then be defined as a function of the difference in temperature between the two nodes, with a cost that increases as the temperature difference increases. The `setCosts` function is then used to apply this cost function to the graph based on the temperature values assigned to each node.

```{r}
exp.cost <-
  exp.cost <- function(x1, x2, cost.coeff) {
    exp(-abs(x1 - x2) * cost.coeff)
  }

# create a set of node costs for temperature
land_graph <- dropDeadEdges(full_graph, thres = 9)
land_graph@nodes.attr$temp <- runif(n = 25821)
temperature_graph <-
  setCosts(
    land_graph,
    node.values = land_graph@nodes.attr$temp,
    method = "function",
    FUN = exp.cost,
    cost.coeff = 1
  )

plot(temperature_graph, edge = TRUE)
plotEdges(temperature_graph)
```

Now we want a combined cost that both captures the temperature as well as the terrain. For this we can use the `combineCosts` function, which allows to combine costs from two different graphs using a specified method (sum, product, or a custom function). Here we will use the `prod` method to combine the costs from the `temperature_graph` and the `land_graph`, which multiplies the costs from both graphs to create a new cost that reflects both the temperature and terrain constraints on movement.

```{r}
combine_costs_graph <- combineCosts(temperature_graph, land_graph, method = "prod")

plot(combine_costs_graph, edge = TRUE)
plotEdges(combine_costs_graph)
title("Combined costs graph")
```


## Extracting information from GIS shapefiles

In the previous sections, we extracted environmental information from raster layers and assigned it to graph nodes using `assignRasterPoints.` Another way *geoGraph* can serve as an interface between geographic information system (GIS) layers and geographic data is the function `extractFromLayer`.
*geoGraph* uses `sf` objects to
represent geographic objects such as points and polygons. By default, *geoGraph*
uses the package *rnaturalearth* to provide continent and country outlines, but
it is possible also to load custom GIS shapefiles with `sf::st_read()`.

We start by loading country outlines for the whole world. Note that we
turn off spherical trigonometry functions with `sf::sf_use_s2(FALSE)`, as the
`naturalearth` dataset is not compatible
with that functionality.
```{r }
library(sf)
sf_use_s2(FALSE)
world.countries <- rnaturalearth::ne_countries(
  scale = "medium",
  returnclass = "sf"
)
```

Let us assume that we are interested in add continent and country information to the
`full_graph` object.
```{r }
newGraph <- extractFromLayer(full_graph,
  layer = world.countries,
  attr = c("continent", "name")
)
summary(getNodesAttr(newGraph))
```

The new object `newGraph` is a `gGraph` which now includes, for each node of the grid, the
corresponding continent and country retrieved from the GIS layer. Note that `extractFromLayer` can extract information to other types of objects than `gGraph` (see `?extractFromLayer`)

We can use the newly acquired information for plotting `newGraph`, by defining new color rules:
```{r fig=TRUE}
temp <- unique(getNodesAttr(newGraph)$"name")
col <- c("transparent", rainbow(length(temp) - 1))
colMat <- data.frame(name = temp, color = col)
head(colMat)
tail(colMat)
plot(newGraph, col.rules = colMat, reset = TRUE)
```

This information could in turn be used to define costs for traveling on the grid.
For instance, one could import habitat descriptors from a GIS, use these values to formulate a habitat model, and derive costs for dispersal on the grid.

### Look at connectivity between two polygons

```{r}
test <- polygonBetween(newGraph, layer = "name", "Andorra", "Portugal", outline = FALSE)
plot(newGraph, col = NA, reset = TRUE)
plot(test)
```


